---
title: "The Great British Bake-Off"
date: last-modified
date-format: "MMMM D, YYYY"
author: 
    - name: Kaatje Matthews-vanKoetsveld
      affil-id: 1
    - name: Alie Hall
      affil-id: 1
    - name: Olivia Kurani
      affil-id: 1
    - name: Hailey Ernest
      affil-id: 1
affiliations:
    - id: 1
      name: Department of Statistics, Cal Poly - SLO
abstract: "Any show’s success is fully reliant on its viewership and audience response. Determining the main factors that relate to an increase or decrease in viewers can be immensely beneficial for a show to sustain its profits and progress. Throughout each episode of the series in ‘The Great British Bake Off’, there are varying viewership trends. Some of the trends that seemed most critical to the success of the show was the viewership for Final versus Non-Finale episodes as well as the overall viewership for series 8 and after. Our analyses show evidence that the finale episode of each series has significantly more viewers than any other episode, with non-finale episodes having between 7.75 to 9.09 million viewers and finale episodes having 7.27 to 12.51 million viewers. During series 8 of ‘The Great British Bake Off’, the popular cooking underwent a few production changes that ultimately had notable consequences on the viewership of that series and the ones to follow. Our analysis of series 8, 9, and 10 compared to the earlier seasons shows that there is a considerable decline in viewership compared to the earlier series. These results can give producers of the show an idea of what will best draw the attention of their audience and make adjustments to up the viewership for ‘The Great British Bake Off’.
"
format: 
  pdf:
    number-sections: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
execute: 
  warning: false
  message: false
  echo: false
bibliography: references.bib
---

```{r}
#| label: setup
library(tidyverse)
library(here)
library(kableExtra)
library(broom)


ratings <- read_csv("ratings.csv")

```

```{r}
#| label: data-cleaning


## ADDING INDICATOR VARIABLE
finale_episodes <- ratings|>
  group_by(series)|>
  slice_max(order_by = episode)|>
  ungroup()|>
  mutate(
    finale = TRUE
  )

non_finale_episodes <- ratings |>
  anti_join(finale_episodes)|>
  mutate(
    finale = FALSE
  )
ratings_new <- non_finale_episodes |>
  full_join(finale_episodes)

## Summary Stats for Q1
non<- non_finale_episodes|>
  summarise(
    type = "Non Finale Episodes",
    mean = round(mean(viewers_7day), digits = 3),
    sd = round(sd(viewers_7day), digits = 3),
    median = round(median(viewers_7day), digits = 3),
    q1 = round(quantile(viewers_7day,probs = .25), digits = 3),
    q3 = round(quantile(viewers_7day,probs = .75), digits = 3)
    
  )

fin <- finale_episodes|>
  summarise(
    type = "Finale Episodes",
    mean = round(mean(viewers_7day), digits = 3),
    sd = round(sd(viewers_7day), digits = 3),
    median = round(median(viewers_7day), digits = 3),
    q1 = round(quantile(viewers_7day,probs = .25), digits = 3),
    q3 = round(quantile(viewers_7day,probs = .75), digits = 3)
  )


## Ratings subset for linear models
ratings_subset <- ratings|>
  filter(series <8)






```

## Introduction {#intro}

Season after season, “The Great British Bake Off” is loved by its fans – but are all series and episodes of the show created equal? Or is there a secret ingredient to the show’s success? “The Great British Bake Off” (GBBO) is a popular baking show that has run from 2010 to present, and features ten to twelve amateur bakers that hail from the Great Britain area. Each episode consists of three different challenges: the signature challenge, the technical challenge, and the showstopper challenge. Each episode, one baker is eliminated until there are three bakers left and then one series winner is chosen, the Star Baker. 
We examined two different trends within viewership of this show. The first investigates if there is a statistically significant difference in finale viewership compared to other episodes. The second discusses the series 8 changes within production of the show and consequences on viewership. This report will include an overview of the statistical methods used to analyze the significance of our research questions, then give an insight on the results of these tests and how the results can be used in the future of “The Great British Bake Off”.

## Data and Methods

We sourced our data on The Great British Bakeoff from TidyTuesday [@tidy], an online community that posts weekly datasets to a public repository for people to share their visualizations and analyses. Included in The Great British Bakeoff package [@bakepack], posted on October 25, 2022, are four datasets, consisting of data from Wikipedia and PBS collected by Alison Hill, Chester Ismay, and Richard Iannone. The data includes information on individual episode challenges, the bakers themselves, the episode ratings, and further information on the episodes. This project only utilizes the "Bakers'' and the "Ratings'' datasets. All of our data analysis will be done in R Version 4.3.2 [@RVersion] The Bakers dataset includes each of the 120 bakers that have appeared on the show from seasons 1-10 (2010-2019) as an observation and contains some basic personal information such as age and occupation, and multiple other quantitative variables regarding their baking challenge scores from their season. The Ratings dataset has each of the 94 aired episodes from seasons 1-10 as observations and records the episode's viewers in the millions for the first 7 days and the first 28 days. Specifically, we are interested in the weekly episode viewership for each episode.

### Data Cleaning and Manipulation

Although we did not truly manipulate any of our variables, we did add one very important indicator variable. This indicator variable was used to determine if an episode was a finale episode or not, which we determined by grouping our data by series, arranging by episode number and then mutating to say finale = TRUE.

In addition, at various points within our analysis, we subsetted different series of our data. To create our reduced linear model, for example, we subsetted the data to only include data from series 1-7.

### Statistical Analysis

We used many different methods of analysis. Firstly, this included a t-test to determine if there was a statistically significant difference in viewership between finale and non-finale episodes. For this t-test, we examined a p-value and a confidence interval of the difference in means. Next, we applied two linear models to our data, one to a subset of our data from the first seven series, and a second one onto the full data. As part of this analysis, we looked at the comparative r-squared values for both of the linear models, and the p-value from a model utility test. We also compared the coefficients for the intercepts and slopes of the models.

## Results {#results}

```{r}
#| label: data-analysis
# you might choose to run your analysis here or in a separate script. Save your output for later.

## T-TEST
t.test(formula = x ~ y, x = ratings_new$viewers_7day, y = ratings_new$finale, alternative = "greater")


## LINEAR MODELS
bakeoff_early_fit <- lm(viewers_7day ~ series, data = ratings_subset)

bakeoff_fit <- lm(viewers_7day ~ series, data = ratings)

```

```{r}
#| label: tbl-finale-viewership-summary
#| tbl-cap: "This table displays summary statisitcs for non-finale episodes and finale episodes. These summary statistics are as follows: mean, standard deviation, median, quartile 1 and quartile 3."
#| tbl-pos: "H"


non |>
  full_join(fin)|>
  kable(
    format = "pipe",
    align = "l",
    caption = "Summary Stats for Non Finale and Finale Episodes"
  )
```

```{r}
#| label: fig-finale-viewership-plot
#| fig-cap: "This graph displays the viewership by series of the Great British Bake Off, with blue dots representing the viewership for non-finale episodes and yellow triangles viewership of finale episodes, with finale episodes having consistantly higher viewership than non-finale episodes."
#| fig-alt: "Your figure alternative text."
#| fig-align: center
#| fig-pos: "H"

ratings_new|>
  ggplot(mapping = aes(x = factor(series)))+
  geom_jitter(mapping = aes(y = viewers_7day, color = finale, shape = finale),
              width = 0.2, height = 0, weight = 4, show.legend = FALSE)+
  theme_bw()+
  annotate(
    "text", x = 6.5, y = 6, label = "Finale episodes marked with\nyellow triangles. Non-finale\nepisodes are blue circles."
  )+
  labs(
    x = "Series",
    y = "Viewership (in millions of views)", #look at this later alie and see if you still want a y-axis label
    subtitle = "Finale Episodes see larger viewership over 7 days than Non-Finale Episodes",
    title = "Tuning In When There's No Next Week: Finales tend to see \nHigher Viewership"
  )+
  scale_color_manual(values = c("steelblue", "goldenrod3"))

```

```{r}
reduced <- glance(bakeoff_early_fit)|>
  select(r.squared, p.value)|>
  mutate(
    type = "reduced",
    r.squared = round(r.squared, digits = 3)

  )
```

```{r}
#| label: tbl-linear-model-comp-and-summary2
#| tbl-cap: "This table displays the R-Squared Values and model utility Test p-values for our respective linear models."
#| tbl-pos: "H"
  
glance(bakeoff_fit)|>
  select(r.squared, p.value)|>
  mutate(
    type = "full",
    r.squared = round(r.squared, digits = 3)
  )|>
  full_join(reduced)|>
  kable(
    caption = "blah blah",
    
  )

```

```{r}
#| label: fig-linear-model-viewership-plot
#| fig-cap: "Graph of viewership for every episode with two lines projected onto it. Blue line is projected onto the graph showing the subsetted linear model. Yellow line is projected onto the line showing the full linear model. This graph shows how the subsetting linear model fits the first seven series extremely well, showing that series 8, 9, and 10 have a different trend."
#| fig-alt: "The figure has black open dots that represent the viewership of each individual episode. There is an increasing trend of viewership from series 1 to 7, which drops during series 8, adn stays consisent during series 9 and 10. In addition, there are two lines on the graph, "
#| fig-align: center
#| fig-pos: "H"

ratings |>
  ggplot(mapping = aes(x = factor(series), y = viewers_7day))+
  geom_jitter(width = 0.2, height = 0, shape = 1)+
  geom_abline(
    slope = 1.97278, intercept = -0.08880, col = "steelblue", linetype = "dashed", size = 1
              )+
  geom_abline(
    slope = 0.76636, intercept = 4.16083, col = "goldenrod3", size = 1
    )+
  theme_bw()+
  labs(
    x = "Series",
    y = "Viewership (in millions of views)",
    subtitle = "Note: Our Linear Models have been projected over this graph for contrast",
    title = "Viewership after 7 days of Episode Release by Series with Linear Models"
  )
  
```

## Discussion {#discussion}

Our results from our t-test p-value indicate there is statistically significant evidence that the mean viewership 7 days after an episode airs is different for finale, 9.89 million viewers, and non-finale episodes, 8.42 million viewers. The t-test with 95% confidence that we conducted provided confidence intervals for the mean viewership of the finale and non-finale episodes, (7.27, 12.51) and ( 7.75, 9.09) respectively. However, there was an overlap in these two intervals which indicates that there is not a significant difference in the mean number of viewers 7 days after the episode aired. We can conclude that there tends to be a correlation but not causation between mean viewership of finale and non-finale episodes. Although we conducted out tests using data from every episode of “the Great British Bake Off”, in order for us to draw conclusions about causation, we would need to be able to replicate each episode of each series. This process of replication would be extremely difficult because there would need to be subject blindness, but a large population is already aware of “The Great British Bake Off”. Despite not being able to make conclusions about causation, we speculate that this increase in viewership associated with Finale episodes has to do with the content and stakes of these specific episodes. Each Finale episode of the series, the series winner is determined after the final three bakers compete for the title of Star Baker. It is likely that this episode has a larger audience because more people are curious to see who wins the whole series rather than who is eliminated during the preceding, ordinary episodes.
 	The contrast in viewership in series 8 and beyond compared to preceding series is also a critical detail that producers of ‘The Great British Bake Off’ should take into consideration. Our models provide strong evidence that there was a notable delince in viewership of series 8 and beyond. Similar to the difference in Finale and Non-Finale episodes, we cannot conclude causation here, but we can speculate that this decline in viewers is associated with the change of hosts and judges that occurred for series 8. Given that the producers make the decision of who the MC’s and judges of the show should be, they should be aware of the consequence these decisions have on viewership. This statistically significant difference in viewership in these series can be translated to a more practical significance as well. If viewership continues to stay at these decreased levels, the producers of the show need to know how much this will affect the show’s revenue. Because this is an analysis based on observations, we cannot fully conclude causation between the production changes and decline in viewership, however we can see that there is a strong correlation between the two. This limits us from fully crediting the series 8 production changes with the decline in viewership which means we can not recommend reversing those changes in order to bring viewership up to its previous numbers. 
	It is important to note that we had census-like data because our datasets included information on every episode from all the series in “The Great British Bake Off” to date. Having access to all of this information greatly supports our conclusions and speculations because it allows us to generalize for future series. However, more episodes have been released since this time so updated data could lead to slightly different conclusions or even more research questions. 


## Conclusion {#conclusion}

We explored the differences in viewership for Finale and Non-Finale episodes of “The Great British Bake Off”, as well as the production changes of series 8’s effect on the viewership from there on out. In order to answer these questions, we utilized t-tests and linear models to get a better sense of the data and their statistical significance. All in all, our analyses have shown that finale episodes of series have a significantly higher viewership than other episodes, and that since the production changes during series 8, the overall viewership of ‘The Great British Bake Off’ has declined. These differences in viewership have real consequences for the overall revenue of the show and how budgets are created for future series. Producers can take extra care when planning finale episodes, since that is when the largest audience is watching. Additionally, producers can try to negotiate for better airing times in an effort to bring back up viewership throughout the series. In the future, it would be intriguing to investigate the ratings based on seasons or baker backgrounds. Researching the baker's background affects the show’s success will provide insight into the show's winners.


## References {.unnumbered}
